Measurement and Probability in Fuzzy Quantum Theories

### 2016-11-26 proposal/prop.tex
p

We propose to revisit quantum mechanics, quantum information, and
quantum computation from this resource-aware perspective. Our initial
result in that domain showed how subtle the issues can be~\cite{usat}:
a straightforward replacement of the complex numbers by a finite field
yields a variant of quantum mechanics in which computationally hard
problems like UNIQUE-SAT (which decides whether a given Boolean
formula is unsatisfiable or has exactly one satisfying assignment) can
be deterministically solved in constant time. To eliminate such
unrealistic theories requires delicate analysis of the structure of
the Hilbert space, the process of observation, and the notion of
probability teasing apart their reliance on the infinitely precise
real numbers~\cite{geometry2013,DQT2014}. In this proposal, our aim is
to shift focus from the infinitely-specified but not directly
observable quantum states, to observable measurable properties of
quantum systems and their probabilities. Furthermore, we insist that
our theories of measurement and probability only refer to finitely
communicable evidence within feasible computational bounds. It follows
that states, observations, and probabilities all become ``fuzzy'',
i.e., specified by intervals of confidence that can only increase in
precision if the available resources increase proportionally. Our
notion of ``fuzzy quantum mechanics'' is related to existing
work~\cite{GranikCaulfield1996,Pykacz2013,SNL2009,Gudder2005,aerts1993physical}
but, as will be explained in more detail, is distinguished by its
computational character.

@@

In a series of breakthrough papers that led to his Turing
Award~\cite{Milner:1993:EIT:151233.151240}, Milner argued that this
``closed system'' view of computation is too limiting as it cannot
adequately describe the increasingly pervasive \emph{interactive
  computations}. Milner explained that ``concurrency requires a new
conceptual framework'' and ``not merely an extension of the repertoire
of entities and constructions which explain sequential computing.'' In
his Turing Award lecture, he gives the following simple but compelling
example. Consider the following two programs:
\begin{verbatim}
Program P1: x := 1; x := x + 1 
Program P2: x := 2
\end{verbatim}
As closed systems, these two programs define the same relation, namely
the one that maps an input memory to an updated memory in which the
location named \verb|x| is assigned the value \verb|2|. An observer
with the ability of inquire about the value of \verb|x| at the end of
the computation cannot distinguish the two programs. But place the
programs in an environment which may involve interactions and the
equivalence is lost. Phrased in physical terms, we give the observer the
additional ability to access the location \verb|x| \emph{during} the
computation, not just to observe its final value. In computing terms,
we place each of the two programs \verb|P1| and \verb|P2| concurrently
with the program:
\begin{verbatim}
Program Q: x := 3
\end{verbatim}
In the combined system of \verb|P1| and \verb|Q| running concurrently,
the location \verb|x| may end up containing \verb|2|, \verb|3|, or
\verb|4|. But in the combined system of \verb|P2| and \verb|Q| running
concurrently, the location~\verb|x| can only end up containing
\verb|2| or \verb|3|. The lesson, phrased in physics terms, is that it
is not sensible to think of the properties of a system (like \verb|P1|
or \verb|P2|) independently of the observer (\verb|Q| in this
case). Indeed the observation \verb|x=4| cannot said to have an
existence independently from the interaction between \verb|P1| and
\verb|Q|: it really emerges or is created as part of the observation
process and is not a pre-existing ``element of reality''.

This phenomenon is hardly a surprise to computer scientists or
physicists but embracing it fully does have far-reaching
consequences. In computer science, the big shift was to abandon the
idea that a computation has inherent internal meaning, e.g.,
representing a pure mathematical function. Instead \emph{the meaning
  of a computational process is its observable behavior.} It is not
too difficult to imagine that, in the presence of multiple interacting
observers, the behavior observed by just one agent would appear to
involve hidden variables or action at a distance. The well-established
theory of observation of concurrent processes and their
equivalences~\cite{Hennessy1989} relies on neither.

@@

This is the most developed part of our current investigation and is
presented in greater detail than the other proposed activities. As
argued in the previous sections, given fixed finite resources, it is
only possible to estimate the quantum probabilities within an interval
of confidence. It is therefore natural to propose the notion of a
``quantum interval-valued probability measure'' that combines the
definitions of conventional quantum probability measures with
classical interval-probabilities.

@@

\paragraph*{Research Question.} Investigate the status of the theorems
of Gleason, Bell, and Kochen-Specker for quantum interval-valued
probability measures. Revisit the Meyer-Mermin debate in this
mathematical framework.

@@

@@

@@

>>>>

We propose to revisit quantum mechanics, quantum information, and
quantum computation from this resource-aware perspective. Our initial
results in that domain showed how subtle the issues can be~\cite{usat}:
a straightforward replacement of the complex numbers by a finite field
yields a variant of quantum mechanics in which computationally hard
problems like UNIQUE-SAT (which decides whether a given Boolean
formula is unsatisfiable or has exactly one satisfying assignment) can
be deterministically solved in constant time. To eliminate such
unrealistic theories requires delicate analysis of the structure of
the Hilbert space, the process of observation, and the notion of
probability teasing apart their reliance on the infinitely precise
real numbers~\cite{geometry2013,DQT2014}. In this proposal, our aim is
to shift focus from the infinitely-specified but not directly
observable quantum states, to observable measurable properties of
quantum systems and their probabilities. Furthermore, we insist that
our theories of measurement and probability only refer to finitely
communicable evidence within feasible computational bounds. It follows
that states, observations, and probabilities all become ``fuzzy'',
i.e., specified by intervals of confidence that can only increase in
precision if the available resources increase proportionally. Our
notion of ``fuzzy quantum mechanics'' is related to existing
work~\cite{GranikCaulfield1996,Pykacz2013,SNL2009,Gudder2005,aerts1993physical}
but, as will be explained in more detail, is distinguished by its
computational character.

@@

In a series of breakthrough papers that led to his Turing
Award~\cite{Milner:1993:EIT:151233.151240}, Milner argued that this
``closed system'' view of computation is too limiting as it cannot
adequately describe the increasingly pervasive \emph{interactive
  computations}. Milner explained that ``concurrency requires a new
conceptual framework'' and ``not merely an extension of the repertoire
of entities and constructions which explain sequential computing.'' In
his Turing Award lecture, he gives the following simple but compelling
example. Consider the following two programs:
\begin{verbatim}
Program P1: x := 1; x := x + 1 
Program P2: x := 2
\end{verbatim}
As closed systems, these two programs define the same relation, namely
the one that maps an input memory to an updated memory in which the
location named \verb|x| is assigned the value \verb|2|. An observer
with the ability of inquire about the value of \verb|x| at the end of
the computation cannot distinguish the two programs. But place the
programs in an environment which may involve interactions and the
equivalence is lost. Phrased in physical terms, we give the observer the
additional ability to access the location \verb|x| \emph{during} the
computation, not just to observe its final value. In computing terms,
we place each of the two programs \verb|P1| and \verb|P2| concurrently
with the program:
\begin{verbatim}
Program Q: x := 3
\end{verbatim}
In the combined system of \verb|P1| and \verb|Q| running concurrently,
the location \verb|x| may end up containing \verb|2|, \verb|3|, or
\verb|4|. But in the combined system of \verb|P2| and \verb|Q| running
concurrently, the location~\verb|x| can only end up containing
\verb|2| or \verb|3|. The lesson, phrased in physics terms, is that it
is not sensible to think of the properties of a system (like \verb|P1|
or \verb|P2|) independently of the observer (\verb|Q| in this
case). Indeed the observation \verb|x=4| cannot be said to have an
existence independently from the interaction between \verb|P1| and
\verb|Q|: it really emerges or is created as part of the observation
process and is not a pre-existing ``element of reality''.

This phenomenon is hardly a surprise to computer scientists or
physicists but embracing it fully does have far-reaching
consequences. In computer science, the big shift was to abandon the
idea that a computation has inherent internal meaning, e.g.,
representing a pure mathematical function. Instead \emph{the meaning
  of a computational process is its observable behavior.} In this
computational analogue to the theory of controlled open quantum
systems, it is not too difficult to imagine that, in the presence of
multiple interacting observers, the behavior observed by just one
agent would appear to involve hidden variables or action at a
distance. The well-established theory of observation of concurrent
processes and their equivalences~\cite{Hennessy1989} relies on
neither.

@@

This is the most developed part of our current investigation and is
presented in greater detail than the other proposed activities. It
relies on current unpublished results. As argued in the previous
sections, given fixed finite resources, it is only possible to
estimate the quantum probabilities within an interval of
confidence. It is therefore natural to propose the notion of a
``quantum interval-valued probability measure'' that combines the
definitions of conventional quantum probability measures with
classical interval-probabilities.

@@

\paragraph*{Research Question.} Investigate the status of the theorems
of Gleason, Bell, and Kochen-Specker for quantum interval-valued
probability measures. Revisit the Meyer-Mermin debate and its
generalizations in this mathematical framework.

@@

The nature of quantum mechanics realized in terms of discrete fields is
amenable to direct computational exploration for reasonable field sizes. We
have established a set of utility libraries in Mathematica that supports a
wide variety of such numerical experiments, with Fig.~\ref{fig_qm} being an
example of the typical output.

Concurrently to this effort, we are also building on our previous
work~\cite{qhaskell,DBLP:journals/entcs/AltenkirchGVS07,DBLP:journals/entcs/Vi\
zzottoCS08,DBLP:conf/wollic/VizzottoBS09,DBLP:conf/sbmf/VizzottoLS09,1166040}
to build semantically well-founded implementations of the computational
models investigated in this proposal.  The aim of these implementations is
two-fold: (i) to provide an experimental testbed that complements the
Mathematica libraries in which discrete quantum algorithms can be simulated;
and (ii) to establish connections between discrete quantum programming
concepts and classical programming concepts.
 We would also like to simulate physical phenomena by using DQCq
 \cite{OrtizEtAl2001,SommEtAl2002}.

\paragraph*{Proposed Activities.}
\begin{itemize}
\item We plan to make our software available for educational and
  research purposes.
\end{itemize}

@@

\noindent (A) {\em Integration of research and education:} Courses on
advanced computational models and quantum information will be further
developed at Indiana University to communicate the basic concepts and tools
discussed in this proposal. Dr. Ortiz taught courses on ``Quantum information   
and Computation'' and Dr. Sabry has taught courses on ``Quantum Computing''
for graduate students.  These courses were highly appreciated by the students
as noted by their evaluations in making the material extremely accessible and
interesting.  Dr.\ Hanson has taught courses on scientific visualization and
mathematical modeling with emphasis on high-dimensional mathematics.  His
numerous Siggraph course presentations are embodied in his 2006 book {\it       
Visualizing Quaternions\/}, which remains the definitive treatment of the
subject for computer science and data analysis applications and has sold
thousands of copies. As a part of Dr.\ Hanson's commitment to exposing and
explaining the elements of higher mathematics to the public, he has recently
led the design of a free iPhone App, {\it 4Dice\/}, that for the first time
makes available an interactive treatment of a correctly imaged simulation of
the appearance of a hypercube, realizing interactively his YouTube animation
of the same name.  This framework is now being generalized for a broad family
of iPhone Apps to permit the hands-on study of high-dimensional mathematics.
The PIs plan to engage high school students or undergraduates in the
implementation of some of the ideas and the exploration of finite geometry.
The PIs have a long standing commitment to K-12 science and math education
(see also (C) below). The collaborative group efforts will provide students,
as well as the PIs with a unique learning experience. The broadened
perspectives gained through those efforts will contribute to the PIs
effectiveness as teachers and mentors. They also will enable to better
prepare students and postdocs for productive careers in an increasingly
competitive global economy. Students and postdocs are encouraged to pursue
collaborative research leading to a broader scientific program. This is
highlighted by the PIs students' close contact with LANL, the Perimeter
Institute, and the Institute for Quantum Computing at Waterloo.

\medskip
\noindent (B) {\em Broadening the participation of underrepresented groups:}
Indiana University is engaged in attracting women and minorities in all
ranks. Several programs are aimed at specifically attracting outstanding
students into physics and computer science, and to expanding the pool of
candidates for fellowships. Targeted recruitment of minority graduate
students is facilitated by the Falicov Fellowship established in 2006. In
recent years, the School of Informatics and Computing has been fortunate to
hire several women and minorities for faculty positions, and we have been
working toward increasing the number of women and minorities among our
student population. To assist with this process, the PIs will encourage women
and minorities to apply for graduate positions as well as undergraduate
positions that may be made available by the grant and REU
supplements. Dr. Ortiz has minority students and Dr. Sabry has graduated a
female Ph.D. in 2005 and is currently assisting on the Ph.D. committees of
several others. Efforts will be made to broaden diversity.

\medskip
\noindent (C) {\em Enhancing infrastructure and outreach:} Dr. Ortiz has
pursued broad outreach by giving talks on astronomy/astrophysics in the local
schools at Bloomington, Indiana, and by an interview in the public radio on
quantum information. Dr. Ortiz is also guiding a high school student in work
on interferometry in two and three slit experiments. Dr. Ortiz also guides
undergraduate research. Dr.\ Hanson plans to add interactive exploration of
the geometry of quantum computing (see Fig.~\ref{fig_qm}) to the suite of
educational iPhone Apps currently being developed. All source code produced
by this project will be made freely available via the web, and the results
will be disseminated via technical reports, conference papers, and journal
articles.  The software and concepts we have developed in the past have been
employed by a wide variety of commercial and noncommercial users.  Thus, we
expect that the successful completion of the proposed research will
contribute to the understanding of quantum and semi quantum computation. The
educational benefits that will flow from this proposal will extend beyond the
students of the participating research groups. The knowledge gained through
participation in the proposed research will further strengthen the
effectiveness of these outreach activities.

Applying within the university.. NSF funding would help us to get
critical mass; new engineering program; credibility importance
university will fund; new lab entanglement

@@

There was no recent NSF support for Dr. Ortiz and Dr. Hanson. In the
summer of 2011, Dr. Sabry received in collaboration with two of his
colleagues two NSF grants: \emph{Effectful Software Contracts} with
Dr. Amal Ahmed and \emph{Reversible Concurrency} with Dr. Geoffrey
Brown. Both proposals are in their initial stages and no significant
results have appeared yet.
